### 2009-11-22 papers/esop2010/rebuttal.txt

Reviewer 1
==========

Lemma 1 of section 5 states the soundness theorem of the WP calculus. The
soundness of the type system itself is included in the precise formulation of
Theorem 1 (in the technical appendix). Despite the superficial differences,
the soundness theorem (and proof) is very similar to the classical
presentation of effect systems, so we judged it to be of less importance.

Concerning the related works, this paper aims at achieving (almost) as much as
others (such as YNot) without using dependent types or separation logic. Our
aim, stated in the section of future works, is to be able to discharge the
majority of the proof obligations by  off-the-shelve first-order provers. This
would bring a higher degree of automation than existing tools, but we are not
there yet.

We do not share the reviewers' aversion against regions. Regions are employed
in many recent papers (Charguéraud/Pottier ICFP 2008), and once an effect
system is used, they are a very simple way to improve the precision of the
system (compared to using program names).

We don't understand the "suspicion" expressed by the reviewer. We believe that
there are basically the following possibilities to deal with aliasing:
  1) separation logic (the systems cited by the reviewer use it)
  2) a very powerful effect system
     (again Charguéraud/Pottier ICFP 2008, for example)
  3) separation analyses (see for example Hubert/Marché [11] and Bornat [4])
  4) pointer (dis-)equalities
We do not want to use separation logic, and we need time to integrate ideas
from the domains 2) and 3) (stated in the future works). The remaining
solution is 4), which is possible in our system, but we consider it to be not
very practical for the difficult case of shared mutable data structures. We
therefore decided to deal with alias-free programs first. But there is no
inherent incapacity of effect systems to deal with aliasing programs.

We think that the paragraph "It would have been a somewhat different
matter ..." does not do justice to our paper. Sections 3 and 4 present a type
system and a wp calculus which *allow* programs with aliasing. The current
prototype implementation treats alias-free programs by default, but to enable
aliasing there is just a simple check in the type system to be removed
(Section 5) and the simplifications enabled by the absence of aliasing
omitted.

Reviewer 2
==========

Yes, distinguishing read/write effects would be a very valuable extension, as
it greatly improve the practical usage of such a system. We are aware of the
importance of such an extension, but haven't worked out the details.

p12 - "The second form of aliasing" - No, this is not about alpha-conversion.
In the usual region calculus and in [24] (Talpin/Jouvelot), a region may
contain more than one reference. You seem to understand the notion of region
in the same way, see your comment about "effects" and "regions" before.
Different references which lay in the same region are aliased in this way -
they cannot be distinguished by the type system. The second restriction of
section 5 disallows allocating twice in the same region, thus eliminating this
form of aliasing. Maybe calling this aliasing is confusing because this notion
usually refers only to the first form in the literature. However, both are
really dual.

Our contribution with respect to Why / Spec# is of cours the support for
higher-order programs; both tools only address first-order programs.

Reviewer 3
==========

Detail a bit more what our concrete contributions are

Concerning aliasing: While reasoning about programs containing aliasing is
possible in our system, we do not claim to present a practical solution for
this problem in this paper. Our system is practical for alias-free programs
and we believe that the examples show this. However, just as separating
analyses other than separation logic are possible in a first-order setting, we
believe this to be true for higher-order programs. These analyses could be
done on top of our system. However, we don't have any experimental data to
support this claim.

Yes, our ML Workshop 2009 paper is mainly about an extended case study using
our tool Who. The citation always appears in connection with the tool our the
case study.

Reviewer 4
===========

Detail a bit more what our concrete contributions are

### 2010-02-18 doc/manual/manual.tex
links in manual

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\cat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}

@@

An \cat{Id} is program identifier or a global type and is defined as: a word
composed of {\tt [A-Za-z0-9']*}, which starts with a letter out ot {\tt
[A-Za-z]}.

A \cat{Region} is the same category.

An \cat{EffectVar} or \cat{Tyvar} is defined as the same category, 
but starting with a {\tt '}.

@@

A \cat{Declaration} is one of the following: 
\begin{itemize}
  \item {\tt axiom} \cat{Id} \cat{Gen} {\tt :} \cat{Term}
  \item {\tt logic} \cat{Id} \cat{Gen} {\tt :} \cat{Type}
  \item {\tt type} \cat{Id} \cat{Gen}  ( = \cat{Type}){\tt ?}
  \item {\tt letregion} \cat{Id}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
    {\tt ,} \cat{Effect} {\tt =} \cat{Allocdef} {\cat{Pre}} {\cat{Post}}
  \item \cat{Letform}
\end{itemize}

A \cat{Letform} is:
\begin{itemize}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Term}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item {\tt let rec} \cat{Gen} '{\tt (}' \cat{Id} {\tt :} \cat{ty} '{\tt )}'
    \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
\end{itemize}

A generalisation \cat{Gen} is:
\begin{itemize}
  \item {\tt [} \cat{Tyvar}* '{\tt |}' \cat{Region} '{\tt |}' \cat{EffectVar}*
    {\tt ]}
\end{itemize}

An argument list \cat{Argslist} is:
\begin{itemize}
  \item ( '{\tt (}' \cat{Id} : \cat{Type} '{\tt )}' )*
\end{itemize}

An \cat{Allocdef} is:
\begin{itemize}
  \item ( allocates \cat{Region} ){\tt ?}
\end{itemize}

A precondition \cat{Pre} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}
\end{itemize}

A postcondition \cat{Post} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}? 
  \item (\cat{Id} {\tt :} )? \cat{Term}
\end{itemize}

A \cat{Type} is one of the following
\begin{itemize}
  \item {\tt int}
  \item {\tt prop}
  \item \cat{Tyvar}
  \item '{\tt (}' \cat{Type} '{\tt )}'
  \item \cat{Id} \cat{inst}?
  \item \cat{Type} \cat{Id}
  \item '{\tt (}' commasep(\cat{Type}) '{\tt )}' \cat{Id}
  \item \cat{Type} {\tt ->} \cat{Type}
  \item \cat{Type} '{\tt *}' \cat{Type}
  \item \cat{Type -}\cat{Effect} \cat{Type}
  \item '{\tt <}' \cat{Effect} '{\tt >}'
  \item {\tt [[} \cat{Type} {\tt ]]}
  \item {\tt ref} '{\tt (}'\cat{Region} \cat{Type}'{\tt )}'
\end{itemize}

An \cat{Effect} is: 
\begin{itemize}
  \item {\tt \{} ( \cat{Id} {\tt |} \cat{EffectVar} )* {\tt \}}
\end{itemize}

An instantiation \cat{inst} is:
\begin{itemize}
  \item {\tt [} commasep*(\cat{Type}) '{\tt |}' \cat{Region} '{\tt |}' \cat{Effect}* {\tt
    ]}
\end{itemize}

A \cat{Term} is one of the following:
\begin{itemize}
  \item '{\tt (}' '{\tt )}'
  \item {\tt !} \cat{Id}
  \item {\tt !!} \cat{Id} ( '{\tt |}' \cat{Term} )
  \item \cat{Id} \cat{inst}?
  \item '{\tt (}' \cat{Term} ( {\tt :} \cat{Type})? '{\tt )}'
  \item \cat{Term} ({\tt \{\{ } \cat{Region} {\tt \}\}} )? \cat{Term} 
  \item \cat{Term} {\tt ;} \cat{Term} 
  \item \cat{Term} \cat{Infix} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item ( {\tt forall} {\tt |} {\tt exists} ) \cat{Argslist} {\tt .} \cat{Term}
  \item {\tt letregion} \cat{Region} {\tt in} \cat{Term}
  \item {\tt if} \cat{Term} {\tt then} \cat{Term} {\tt else} \cat{Term}
  \item \cat{Letform} {\tt in} \cat{Term}
  \item {\tt for} \cat{Id} = \cat{Term} ( {\tt to | downto} ) \cat{Term} {\tt
    do} {\cat{Pre}} \cat{Term} {\tt done}
  \item {\tt [[} \cat{Pre} {\tt ]]} \cat{Term} {\tt [[} \cat{Post} {\tt ]]}
\end{itemize}

>>>>

\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{xcolor}
\definecolor{dablue}{rgb}{0.1,0.1,0.4}

\usepackage[colorlinks=true,pdfborder={0,0,0,0},urlcolor=dablue,
linkcolor=dablue, pdftex]{hyperref}

%\newcommand{\cat}[1]{$<$\textcolor{Aquamarine}{#1}$>$}
\newcommand{\cat}[1]{$<$\hyperref[#1]{#1}$>$}
\newcommand{\catdef}[1]{\label{#1}$<$\textcolor{Aquamarine}{#1}$>$}

@@

\paragraph{}
An \catdef{Id} is program identifier or a global type and is defined as: a word
composed of {\tt [A-Za-z0-9']*}, which starts with a letter out ot {\tt
[A-Za-z]}.

\paragraph{}
A \catdef{Region} is the same category.

\paragraph{}
An \catdef{EffectVar} or \catdef{Tyvar} is defined as the same category, 
but starting with a {\tt '}.

@@

\paragraph{}
A \catdef{Declaration} is one of the following: 
\begin{itemize}
  \item {\tt axiom} \cat{Id} \cat{Gen} {\tt :} \cat{Term}
  \item {\tt logic} \cat{Id} \cat{Gen} {\tt :} \cat{Type}
  \item {\tt type} \cat{Id} \cat{Gen}  ( = \cat{Type}){\tt ?}
  \item {\tt letregion} \cat{Id}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
  \item {\tt parameter} \cat{Id} \cat{Gen} \cat{Argslist} {\tt :} \cat{Type}
    {\tt ,} \cat{Effect} {\tt =} \cat{Allocdef} {\cat{Pre}} {\cat{Post}}
  \item \cat{Letform}
\end{itemize}

\paragraph{}
A \catdef{Letform} is:
\begin{itemize}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Term}
  \item {\tt let} \cat{Id} \cat{Gen} \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item {\tt let rec} \cat{Gen} '{\tt (}' \cat{Id} {\tt :} \cat{ty} '{\tt )}'
    \cat{Argslist} {\tt =} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
\end{itemize}

\paragraph{}
A generalisation \catdef{Gen} is:
\begin{itemize}
  \item {\tt [} \cat{Tyvar}* '{\tt |}' \cat{Region} '{\tt |}' \cat{EffectVar}*
    {\tt ]}
\end{itemize}

\paragraph{}
An argument list \catdef{Argslist} is:
\begin{itemize}
  \item ( '{\tt (}' \cat{Id} : \cat{Type} '{\tt )}' )*
\end{itemize}

\paragraph{}
An \catdef{Allocdef} is:
\begin{itemize}
  \item ( allocates \cat{Region} ){\tt ?}
\end{itemize}

\paragraph{}
A precondition \catdef{Pre} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}
\end{itemize}

\paragraph{}
A postcondition \catdef{Post} is: 
\begin{itemize}
  \item \cat{Term}{\tt ?}? 
  \item (\cat{Id} {\tt :} )? \cat{Term}
\end{itemize}

\paragraph{}
A \catdef{Type} is one of the following
\begin{itemize}
  \item {\tt int}
  \item {\tt prop}
  \item \cat{Tyvar}
  \item '{\tt (}' \cat{Type} '{\tt )}'
  \item \cat{Id} \cat{Inst}?
  \item \cat{Type} \cat{Id}
  \item '{\tt (}' commasep(\cat{Type}) '{\tt )}' \cat{Id}
  \item \cat{Type} {\tt ->} \cat{Type}
  \item \cat{Type} '{\tt *}' \cat{Type}
  \item \cat{Type -}\cat{Effect} \cat{Type}
  \item '{\tt <}' \cat{Effect} '{\tt >}'
  \item {\tt [[} \cat{Type} {\tt ]]}
  \item {\tt ref} '{\tt (}'\cat{Region} \cat{Type}'{\tt )}'
\end{itemize}

\paragraph{}
An \catdef{Effect} is: 
\begin{itemize}
  \item {\tt \{} ( \cat{Id} {\tt |} \cat{EffectVar} )* {\tt \}}
\end{itemize}

\paragraph{}
An instantiation \catdef{Inst} is:
\begin{itemize}
  \item {\tt [} commasep*(\cat{Type}) '{\tt |}' \cat{Region} '{\tt |}' \cat{Effect}* {\tt
    ]}
\end{itemize}

\paragraph{}
A \catdef{Term} is one of the following:
\begin{itemize}
  \item '{\tt (}' '{\tt )}'
  \item {\tt !} \cat{Id}
  \item {\tt !!} \cat{Id} ( '{\tt |}' \cat{Term} )
  \item \cat{Id} \cat{Inst}?
  \item '{\tt (}' \cat{Term} ( {\tt :} \cat{Type})? '{\tt )}'
  \item \cat{Term} ({\tt \{\{ } \cat{Region} {\tt \}\}} )? \cat{Term} 
  \item \cat{Term} {\tt ;} \cat{Term} 
  \item \cat{Term} \cat{Infix} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Term}
  \item fun \cat{Argslist} {\tt ->} \cat{Allocdef} {\cat{Pre}} \cat{Term} {\cat{Post}}
  \item ( {\tt forall} {\tt |} {\tt exists} ) \cat{Argslist} {\tt .} \cat{Term}
  \item {\tt letregion} \cat{Region} {\tt in} \cat{Term}
  \item {\tt if} \cat{Term} {\tt then} \cat{Term} {\tt else} \cat{Term}
  \item \cat{Letform} {\tt in} \cat{Term}
  \item {\tt for} \cat{Id} = \cat{Term} ( {\tt to | downto} ) \cat{Term} {\tt
    do} {\cat{Pre}} \cat{Term} {\tt done}
  \item {\tt [[} \cat{Pre} {\tt ]]} \cat{Term} {\tt [[} \cat{Post} {\tt ]]}
\end{itemize}
